Multiphase implementation on a parallel 2D/3D Lattice Boltzmann solver using GPGPUs

### 2017-08-14 Thesis/Draft.tex
finished intro

\chapter{Literature review}
In this chapter, we will be taken into the broad field of the Lattice Boltzmann method using the CUDA platform. The topic of this thesis is associated with a number of different study areas, such as Computational Fluid Dynamics, Lattice Boltzmann method, High performance computing and GPGPU. We will then present an extensive literature review on these themes in the remaining of this chapter. Firstly, we will look into Computational Fluid Dynamics and the most commonly used approaches to it. Then we will be giving an overview on the Lattice Boltzmann method. Afterwards we will discuss how it is possible to employ parallelisation techniques in scientific computing. Finally we will review some previous works in this area.
\section{Computational Fluid Dynamics}
Computational Fluid Dynamics (CFD) comes from the need to model fluid flows and associated processes.
A wide range of applications come from studying CFD, notably:
\begin{itemize}
\item Aircraft design \cite{aircraft_cfd}
\item Solid particle erosion \cite{partic_erosion_cfd}
\item Wind flow simulation \cite{wind_flow_cfd}
\item Combustion chamber simulations \cite{combust_cfd}
\item Environmental and weather prediction \cite{weather_pred}
\item Automotive and motor sports \cite{america_cup,race_car_wing}
\end{itemize}

@@

\section{High Performance Computing}
As humanity evolves, so too does our desire for expanding previous unobtainable goals. As computer technology kept progressing further and further, we soon realised that some problems simply took too many resources to be completed.\par
Figure \ref{fig:moore} shows the evolution of computing power over the past 120 years. Moore's Law states that ''processor speeds, or overall processing power for computers will double every two years'' \cite{moore}. Note that the last 7 most recent data points are NVIDIA GPUs.

@@

\subsection{Distributed memory}
Following a distributed memory architecture for parallel computing means that each processor will have it's own independent local memory. In these systems, all the work that one process executes remains local to it, without interfering with the address space of all other processors. Hence, to solve meaningful computations, a communication network needs to be established in order for processes to share data with each other. Such is the case of the Message Passing Interface (MPI).\par
This means that this architecture is very scalable: with each processor being added to the system the size of the total memory increases. Also, each processor will be able to access it's own memory rapidly and without interference, reducing the usual constraint of memory access penalties.
\par However, this type of system requires a higher degree of skill from the programmer. The programmer will be the one responsible for most of the details of memory passing between processors and will need to ensure that no race conditions or deadlocks arise from the data communication. Also, whenever data from another processor is needed, the latency of the bandwidth in the network will introduce a heavy time penalty to the computations, as this data will need to be communicated before computations can be performed over it.
\subsection{Shared memory}
In a shared memory architecture, each processor is able to access all memory as global address space. This means that data can be handled seamlessly between processors, making programs easy to read and easy to write. An example of this model can be the Open Multi-Processing (OpenMP) API, which supports shared memory multiprocessing programming in C, C++ and Fortran. In this architecture, processors can operate independently while having access to the same memory resource pool as all other processors. Therefore, changes in the data handled by one process is visible and updated for all others. \par
This means that the programmer can have a user-friendly environment while working on his/her algorithm. All details concerning data flow between processes are abstracted, making this model very beneficial for users with little to no background in parallel programming. Also, the data being shared between processors is fast and uniform, making the penalties of memory access of this shared data less penalizing than in the distributed memory model.
\par However, this architecture presents little memory scalability, since adding more CPUs will increase the traffic of the shared memory path. Also, the programmer will have to pay close attention to memory access so as to prevent memory violation and ensure a correct synchronization of the access to the global adress space.

>>>>

Fluid dynamics is an ever-growing, ever-present part of the modern industry. Almost every field in the industry will eventually have to take fluid dynamics into account when designing and developing their product, from aircrafts, to combustion chambers and even in weather prediction systems. With modernization comes an increase in the amount of techniques available to process these issues, and one of them comes from the use of computers. Instead of spending a high amount of time, effort and, most importantly, money, in building and developing structures such as wind tunnels to study the design of an aircraft, for example, companies can take advantage of Computational Fluid Dynamics and simulate their needs using software. \par
This was traditionally done by solving classic Navier-Stokes equations that treat the fluid as a whole and analyze it's properties. Other ways could rely on the microscopic parts of the fluid and take each into particle into account. Another, great option would be to combine these 2 approaches into a median approach, which is what the Lattice Boltzmann method does, based on a statistical approach to solve the distribution functions associated with the fluid. This method also brings the advantage of it's locality nature. Each node's dependency on it's neighbors is very small, allowing for very few interactions between them. This makes the Lattice Boltzmann method an excellent candidate for parallel computations.\par
At the same time, the method's relative simplicity allows for the modeling of more complex and interesting flows, like Multiphase flows. These flows represent materials in different states, allowing for very interesting interactions like liquid/solid, gas/liquid or even liquid/liquid which will be the main focus of this thesis. These flows extend the applicability of the method to include capillarity problems, paper-making and even fluid management in space. \par
However, to perform a parallel implementation of the methods, one needs hardware on which to run the simulations. Nowadays, we can find multi-processing units in almost every computer, we can even find these in our smart-phones. Large scale simulations, like most used to simulate real-world problems require bigger computation abilities. High Performance Computing aims to enable these types of simulations, and they are usually used in centers specifically created to cover these needs. However, the clusters of computational units used are usually very expensive, but luckily there is a cheaper, viable alternative - GPUs. In 2007, NVIDIA released the CUDA platform, allowing users to take advantage of the processing power of their GPUs to perform General Purpose computing. GPUs offer the ability of using a very large amount of threads to parallelize the work by exploiting their architecture, at the cost of having to understand their complicated architecture and memory management system. \par

\section{Objectives}
This thesis aims to implement Multiphase flows in 2D and 3D into the already existing in-house Lattice Boltzmann solver that has been developed over the past years by several MSc Cranfield students  \cite{maciej_thesis,koleszar_thesis,jozsa_thesis,szoke_thesis}. This implementation will be done in collaboration with a CFD student, Figueroa \cite{antonioThesis}. He will be in charge of the physical aspects of solver, from the development of the models to the validation of the results, while the author of this thesis will be in charge of the software aspect, from implementing the model in the in-house solver, to parallelization and performance improvements. \par
We will first start with the model formulation, where Figueroa will implement said models in Matlab and validate them accordingly. Afterwards, the author will then perform a serial implementation in C into the in-house solver and, after validating the results, parallelize the model using the CUDA platform. After the parallel implementation is validated, we will repeat this process for the 3D model and finally focus on improving the solver's performance. 
\section{Structure}
We will start by introducing the reader into the main areas of the topic in the literature review. There, we can find a description of the methods and applications used in CFD, followed by a an overview of the Lattice Boltzmann methods and Multiphase flows. We then present a brief history on High Performance Computing and the types of parallel programming models, including CUDA. Finally we present the work done by the previous authors of the solver.\par
In the Methodology section, we give some insight into how the Lattice Boltzmann method works and the needs it imposes. We then present the Multiphase model that we will implement in this thesis. Then we follow with a brief discussion on the types of meshes that we will use and we end that chapter with the details of CUDA programming: how it is different from other parallel programming languages, how it's architecture is defined and the various types of memory available for use.\par
In the Results and Discussion chapter we present the work done during this thesis and analyze the results obtained. We start by exposing the in-house solver, how it is built, it's performance and how the code works. Then we present how we developed the Multiphase model into the LBM solver. Afterwards, we discuss the various test cases employed in the model and how they were validated. We then analyze the memory usage of the solver and end the chapter with the performance that our model obtained, as well as the results of parallelization and optimization.\par
Finally, we finish the Conclusion chapter where we perform an overview of the thesis and possibilities for future work in the solver.
\chapter{Literature review}
In this chapter, we will be taken into the broad field of the Lattice Boltzmann method using the CUDA platform. The topic of this thesis is associated with a number of different study areas, such as Computational Fluid Dynamics, Lattice Boltzmann method, High performance computing and GPGPU. We will then present an extensive literature review on these themes in the remaining of this chapter. Firstly, we will look into Computational Fluid Dynamics and the most commonly used approaches to it. Then we will be giving an overview on the Lattice Boltzmann method. Afterwards we will discuss how it is possible to employ parallelization techniques in scientific computing. Finally we will review some previous works in this area.
\section{Computational Fluid Dynamics}
Computational Fluid Dynamics (CFD) comes from the need to model fluid flows and associated processes.
A wide range of applications come from studying CFD, notably:
\item Aircraft design \cite{aircraft_cfd}
\item Solid particle erosion \cite{partic_erosion_cfd}
\item Wind flow simulation \cite{wind_flow_cfd}
\item Combustion chamber simulations \cite{combust_cfd}
\item Environmental and weather prediction \cite{weather_pred}
\item Automotive and motor sports \cite{america_cup,race_car_wing}

@@

\section{High Performance Computing}
As humanity evolves, so too does our desire for expanding previous unobtainable goals. As computer technology kept progressing further and further, we soon realized that some problems simply took too many resources to be completed.\par
Figure \ref{fig:moore} shows the evolution of computing power over the past 120 years. Moore's Law states that ''processor speeds, or overall processing power for computers will double every two years'' \cite{moore}. Note that the last 7 most recent data points are NVIDIA GPUs.

@@

\subsection{Distributed memory}
Following a distributed memory architecture for parallel computing means that each processor will have it's own independent local memory. In these systems, all the work that one process executes remains local to it, without interfering with the address space of all other processors. Hence, to solve meaningful computations, a communication network needs to be established in order for processes to share data with each other. Such is the case of the Message Passing Interface (MPI).\par
This means that this architecture is very scalable: with each processor being added to the system the size of the total memory increases. Also, each processor will be able to access it's own memory rapidly and without interference, reducing the usual constraint of memory access penalties.
\par However, this type of system requires a higher degree of skill from the programmer. The programmer will be the one responsible for most of the details of memory passing between processors and will need to ensure that no race conditions or deadlocks arise from the data communication. Also, whenever data from another processor is needed, the latency of the bandwidth in the network will introduce a heavy time penalty to the computations, as this data will need to be communicated before computations can be performed over it.
\subsection{Shared memory}
In a shared memory architecture, each processor is able to access all memory as global address space. This means that data can be handled seamlessly between processors, making programs easy to read and easy to write. An example of this model can be the Open Multi-Processing (OpenMP) API, which supports shared memory multiprocessing programming in C, C++ and Fortran. In this architecture, processors can operate independently while having access to the same memory resource pool as all other processors. Therefore, changes in the data handled by one process is visible and updated for all others. \par
This means that the programmer can have a user-friendly environment while working on his/her algorithm. All details concerning data flow between processes are abstracted, making this model very beneficial for users with little to no background in parallel programming. Also, the data being shared between processors is fast and uniform, making the penalties of memory access of this shared data less penalizing than in the distributed memory model.
\par However, this architecture presents little memory scalability, since adding more CPUs will increase the traffic of the shared memory path. Also, the programmer will have to pay close attention to memory access so as to prevent memory violation and ensure a correct synchronization of the access to the global address space.
